From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-03-05 paper/paper.Rmd
WIP

To emphasize the underlying trend in the changing roles in which we
see and understand noisy processes, I will also restrict my focus 
to relatively simple models primarily from population ecology context.
Simplicity not only makes examples (in equations and in code) more
tractable but also allows us to focus on aspects that are germane to many
contexts rather than unique to particular complexitites [@Levins1966; @Bartlett1960].
Nevertheless, that complexity matters -- few themes have been better 
emphasized in the theoretical literature [@Bjornstad2001].  Both
the foundational literature and recent research continue to echo the
theme of understanding the impact different real world complexities
have in stochastic dynamics, including the importance
of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989; @Tuljapurkar1997;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g.
@Durrett1994; @Dieckmann2000; @Kerr2002; @Schreiber2010], Individual
heterogeneity [e.g. @Vindenes2015; @Hart2016] or the coexistence
of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007;
@Schreiber2017]; correlated [@Ripa; @Ripa1999; @Petchey, @Schreiber2010;
@Lee2017; @Spanio2017] or periodic [@Coulson2001; @Bjornstad2001;
@Keeling2001] structure in environmental noise, or the interaction
of ecological and evolutionary processes [@Saether1997; @Ozgul2009;
@Lenormand2009; @Vindenes2015; @Schreiber2015].  As such, we will 
rely on both textbooks and recent reviews to provide a proper treatment
of these issues, and focus on broader trends that may be less apparent
to non-specialists.

This review is structured into three sections: Origins of noise, emergent
phenomena, and noise-driven inference.  The first section lays the conceptual
groundwork we will need, while also highlighting a shift to more and more
mechanistically rooted descriptions of noise.  We will see where the common
formulation of 'deterministic skeleton plus noise term' comes from, how it 
is best justified, and when it breaks down.  The second section introduces
"noise the creator," showing examples of ecological phenomena generated by
stochasticity.  These examples will be familiar to most specialists but
illustrate a different way of thinking than held by most ecologists, where
noise is only a nuciance to be filtered or averaged out.  The third and
final section, "noise the informer," turns these examples back-to-front, 
asking what noise can tell us about a system: such as its underlying resilience
or stability, or the approach of a catastophic shift. Examples are fewer
here, and have largely yet to benefit from the introduction of either
the rigorous theorems or more complex models so plentiful in the previous
sections.  Yet the promise of prediction, of early warning signs before
tipping points, have spurred broad interest in such noise-based inference.
This review is a call to both deepen the connection to mechanism and better
formalize this thinking, but also look more broadly into other ways in which
noisy phenomena can help inform and predict underlying processes.

@@

Outside of the primarily theoretical literature on stochastic dynamics
in ecology, noise is viewed largely as a nusiance: something entirely
distinct from the processes we are interested in which invariably keeps
getting in the way.  As we have seen in section on origins, such 
characterizations are misleading: noise arises from the very same processes.
While both empirical and statistical literature have frequently emphasized
techniques to remove noise, a central theme of stochastic processes
have emphasized an almost opposite conclusion: noise cannot be subtracted or
averaged away, since tthe presence of noise can fundamentally alter the
dynamics we observe.
Stochasticity can give rise to a wide array of interesting phenomena
that are often at odds with our intuition. As these examples will
illustrate, such behavior arises through an interaction between
stochastic and non-linear elements of a system, underscoring the
importance of explicitly including stochasticity in ecological
models [@Coulson2004; @Higgins1997].  Importantly, the noise itself
does not have to be large to give rise to these phenomena.  In this
section, I highlight several examples that have shown up frequently
in the literature and prove sufficiently tractable to provide a
simple intuition for the mechanisms involved.

@@

Oscillations arise through the process of stochastic resonance.
The damped oscillations seen in Figure \ref{quasicycles}A are analgous to a child
sitting still on a swing: friction slowly damps the magnitude of
each successive oscillation, while the period of pendulum remains
unchanged.  Driving the pendulum through perturbations timed to
match that natural period can quickly create sustained oscillations
with large magnitudes, but such perfect timing is not required.  In
the model above, white noise effectively drives the pendulum at all
frequencies, though at very low magnitude, since the noise is small
relative to the scale of the dynamics.  However, frequencies that
most closely match the natural period of the damped oscillator are
amplified, resonating with the natural oscillations.  This process
is sufficient to permit even small noise to drive sustained
oscillations in the model indefinitely.  The power spectrum (the
squared norm of the Fourier transform, a measure of what frequencies
are present; see, e.g. @Black2012) provides a convenient way to
visualize the effects of these oscillations.  The power spectrum
for a deterministic limit cycle would create a perfect sinusoidal
oscillation and a resulting power spectrum of a delta spike at that
single frequency.  The power spectrum for the quasi-cycle instead
shows a distribution of frequencies, centered at the the resonant
frequency of the damped oscillator.

Since @Nisbet1976, examples of stochastic resonance have been demonstrated in common
models and implicated in empirical examples of increasing complexity [@Greenman2003].
@Blarer1999 identified quasi-cycles as a possible
explanation for oscillations and periodic outbreaks in pest ecosystems.
@Bjornstad2004 analyzes the potential for such stochastic effects
to drive cyclic behavior in population abundances of Atlantic bluefin
tuna and Atlantic Cod.  @McKane2005, @Alonso2007, and @Black2010 illustrate
the emergence of quasi-cycles in common disease outbreak models, including
the use of power spectrum analysis.  
@Pineda-Krch2007 compare patterns from stochastically generated cycles to
stable limit cycles they resemble to ask how we can discriminate between
these mechanims.  Stochastic resonance highlights
one way in which even small noise can play an important role in
driving large scale patterns where we had previously only sought
deterministic explanations. This example also illustrates the
interplay between nonlinear dynamics and stochasticity that is the
hallmark of stochastically driven phenomena.  A stable node with
linear dynamics, such as an Ornstein-Uhlenbeck process, has no
resonant frequency, and could not produce sustained oscillations
(unless the environmental driver itself had a periodic pattern)
[e.g. @Gardiner2009; @Black2010].

@@

<!--Figure 3: Noisy switch -->
```{r noisy_switch, fig.width=7, fig.height=7, fig.cap="Stochastic switiching in May's model of alternative stable state dynamics.  r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000. \\label{noisy_switch}"}

@@

Because these transitions between alternative stable states appear as
such a dramatic deviation from the previous pattern (as illustrated
in Figure \ref{noisy_switch}A), it is  tempting to assume they
result from some sudden external force or change in the
environment. As the numerical examples clearly illustrate,
these transitions can neverthless be explained entirely by chance
[@Boettiger2012;  @Drake2013].

@@

Transitions from systems with particularly shallow wells have recently
taken on new empirical significance.  Because such transitions become
more likely as the potential well becomes less deep (i.e. the attractor
has become less stable), the onset of such transitions can be intepreted
as an indication of this loss of stability. This phenomenon has been
dubbed "flickering" and suggested as a warning sign that a more
permanent transition to the alternate stable state is approaching [@Brock2010].  
Examples of flickering between alternative
stable states have been suggested in data ranging from geological
timescales at the end of glaciation in the Pleistocene [@Taylor1993]
to ecological scales over a few decades in lake ecosystems [@Wang2012].
Flickering is then our first example of "noise the informer" -- in
which the emergence of a stochastic phenomenon is re-interpreted 
in a way that sheds light on an underlying process.

In examining "noise the creator," we began with a tension between
the intepretation of noise as it is usually presented in the context of
statistical inference: a nusiance to be stripped away, and how it is
viewed by theorist, as a potentially fundamental driver of patterns,
including oscillations, regime shifts, extinction & coexistence.
In this third section, we attempt to re-unite these two largely
divergent views of the role of noise by highlighting how phenomena
from the latter (noisy phenomena) can be used to address the questions
of the former (inferring underlying process from pattern).

Our most provacative example of how noisy phenomena can act to reveal
underlying processes comes from the literature we have just touched upon
in the case of flickering -- a growing set of so-called 'early
warning signs' of critical transitions [e.g. @Scheffer2009].  The fundamental
insight behind all of these signs comes from the exploiting the ability of
equations we encountered in "origins of noise"
to reveal subtle changes in underlying processes before those changes are
visible at the macroscopic (deterministic) level.  This literature focuses
on the phenomenon of a tipping point, in which gradual changes to the environment
show essentially no changes to the behavior of a system (i.e. the abundance of a
population or health of lake ecosystem) followed by a rapid change when the
tipping point is reached that cannot be easily reversed. [@Scheffer2009;
@Scheffer2015].  The fundamental insight of early warning signals has been
to realize that though the average or deterministic dynamics of such a system
do not change significantly prior to the tipping point, the dynamics of the
noise must change in a consistent and predictable manner.

This pattern, known as 'critical slowing down' for reasons that will become 
obvious, was first proposed an an early warning sign for ecological
transitions by @Wissel1984, but has received considerable attention only more recently,
particularly owing to work in empirical systems [@Scheffer2015].
Our analysis in the origins of noise provides sufficient framework to illustrate
this idea completely.  The premise behind the tipping point just described
is that of a fold bifurucation [which I will not detail here, but see @Holling1973; @Scheffer2009; @Strogatz2001].
The key thing we need to know about this bifurcation is that it involves
changes *derivative* of the deterministic skeleton ($\partial_x \left[ b(x) - d(x) \right]_{x=\hat x}$) in
the notation of our example), which can build up for some time without
significant changes to the location of the equilibrium state $\hat x$.
This derivative did not appear in our marcroscopic equation, $\tfrac{\mathrm{d}x}{\mathrm{d}t} = b(x) - d(x)$,
but featured regularly in our equations for the noise, such as the 
expected variance, Eq \eqref{var} and autocorrelation, Eq \eqref{autocor}.
This derivative represents the *stability*  of the equilibrium: how quickly
deviations return to the stationary value.  
The magnitude of both variance and autocorrelation must increase
as the stability ($\partial_x \left[ b(x) - d(x) \right]_{x=\hat x}$) decreases
prior to a critical tranisition.

<!-- Figure 4: Early warning signals -->
```{r tipping, fig.width=7, fig.height=10, fig.cap="Early warning signs of a critical transition in a simulation of increasing serial dilution of budding yeast. Panel A: Data from replicate manipulations by @Dai2015. Panel B: Coefficient of variation for the same experimental data. Panel C: Simulated single time series moel.  Panel D, E: Rolling variance and autocorrelation over simulated data show an increase in advance of the transition.  Simulation details in Appendix A. Warning signal statistics for experimental data not shown, see @Dai2015.\\label{tipping}"}
tipping <- read_csv("../appendixA/tipping.csv")

@@

The trick of the early warning signals approach is to view noise as 
a sequence of countless miniature perturbation experiments.  
Figure \ref{tipping} illustrates early warning indicators in an
experimental manipulation and corrsponding simulated model of
@Dai2012 and @Dai2015.  Though early warning signals have been
reported in a growing range of systems [@Scheffer2015], these
laboratory experiments in population dynamics
they observed in serial dilutions of budding yeast _Saccharomyces
cerevisiae_ on a sucrose medium provide the most convincing 
mechanistic intepretation of the underlying fold bifurcation.
Because the relevant proteins are
located in the extra-cellular matrix rather than inside the cells,
budding yeast break down sucrose collaboratively, creating an Allee
effect below which the population is too small to sustain itself.
Increasing the serial dilution rate slowly over time drives the
stable population towards a fold bifurcation which is accompanied
by the patterns of increasing variance and increasing auto-correlation.
This pattern can be seen empirically in lab populations [@Dai2012; @Dai2015]
and explored numerically in a simulation of the yeast dynamics shown in 
Figure \ref{tipping}. As the mechanistic model of yeast population
dynamics is already quite simple, the simulation shown here (code in Appendix B)
corresponds directly to the model proposed by @Dai2015.

@@

In contrast to the examples we have considered previously, work that extends
this analysis into more complex scenarios of structured populations, interacting species,
autocorrelated noise etc are significantly underdeveloped. Issues of spatial
structure have recieved more attention [e.g. @Guttal2009; @Dakos2011; @Bel2012; @Dai2013]
but still largely idosynchratic.  Significant challenges remain
in distinguishing these patterns from other causes [@Boettiger2013a].
For instance, as we have seen, variance in a system can increase due to
external drivers without changing the auto-correlation, and both variance
and auto-correlation can increase if a system loses stability without
approaching a saddle-node bifurcation, but will not show flickering.
Teasing these issues apart in greater mathematical formalism and under more
complex and realistic models remain largely open problems.

More broadly, examples of early warning signals such as increasing variance,
auto-correllation, and flickering all point to ways in which stochasticty
and noise-driven phenomena can be transformed from something which obscures
underlying processes to something which reveals them.  I believe this is
role could apply far more broadly than in just the context of stability and
tipping points.  The examples from fold bifurcations already suggest
that stochastic processes could play a similar role in identifying
other critical transitions. For instance, stochastic resonance may be 
visible in advance of a Hopf bifurcation, where a stable state gives
way to stable oscillations (limit cycle) of growing magnitude
[e.g. @Fussmann2000; @Nelson2013; @Strogatz2001].

Intepreting stochastic processes as potential new sources of information
has also proven to be a
rich catalyst for bringing these concepts beyond the realm of abstract
theory. A rich array of experimental and observational tests of early
warning signals [_i.e._ @Carpenter2011; @Dai2012; @Clements2016; @Clements2017;
and many others, _e.g._ see @Scheffer2015 review] has served not merely
to illustrate these phenomena can be seen in laboratory and field ecosystems,
but also to spark suggestions of additional possible indicators and surface
a wide array of practical [@Boettiger2013b; @Scheffer2015], statistical
[e.g. @Boettiger2012d; @Boettiger2012], and theoretical [e.g. @Dai2015;
@Hastings2010] challenges to such application.  This creates a two-way dialog
between empirical and theoretical work that does much to advance the field
as a whole.

>>>>

To emphasize the underlying trend in the changing roles in which we
see and understand noisy processes, I will also restrict my focus 
to relatively simple models primarily from population ecology context.
Simplicity not only makes examples (in equations and in code) more
tractable but also allows us to focus on aspects that are germane to many
contexts rather than unique to particular complexities [@Levins1966; @Bartlett1960].
Nevertheless, that complexity matters -- few themes have been better 
emphasized in the theoretical literature [@Bjornstad2001].  Both
the foundational literature and recent research continue to echo the
theme of understanding the impact different real world complexities
have in stochastic dynamics, including the importance
of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989; @Tuljapurkar1997;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g.
@Durrett1994; @Dieckmann2000; @Kerr2002; @Schreiber2010], Individual
heterogeneity [e.g. @Vindenes2015; @Hart2016] or the coexistence
of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007;
@Schreiber2017]; correlated [@Ripa; @Ripa1999; @Petchey, @Schreiber2010;
@Lee2017; @Spanio2017] or periodic [@Coulson2001; @Bjornstad2001;
@Keeling2001] structure in environmental noise, or the interaction
of ecological and evolutionary processes [@Saether1997; @Ozgul2009;
@Lenormand2009; @Vindenes2015; @Schreiber2015].  As such, we will 
rely on both textbooks and recent reviews to provide a proper treatment
of these issues, and focus on broader trends that may be less apparent
to non-specialists.

This review is structured into three sections: Origins of noise, emergent
phenomena, and noise-driven inference.  The first section lays the conceptual
groundwork we will need, while also highlighting a shift to more and more
mechanistically rooted descriptions of noise.  We will see where the common
formulation of 'deterministic skeleton plus noise term' comes from, how it 
is best justified, and when it breaks down.  The second section introduces
"noise the creator," showing examples of ecological phenomena generated by
stochasticity.  These examples will be familiar to most specialists but
illustrate a different way of thinking than held by most ecologists, where
noise is only a nuisance to be filtered or averaged out.  The third and
final section, "noise the informer," turns these examples back-to-front, 
asking what noise can tell us about a system: such as its underlying resilience
or stability, or the approach of a catastrophic shift. Examples are fewer
here, and have largely yet to benefit from the introduction of either
the rigorous theorems or more complex models so plentiful in the previous
sections.  Yet the promise of prediction, of early warning signs before
tipping points, have spurred broad interest in such noise-based inference.
This review is a call to both deepen the connection to mechanism and better
formalize this thinking, but also look more broadly into other ways in which
noisy phenomena can help inform and predict underlying processes.

@@

Outside of the primarily theoretical literature on stochastic dynamics
in ecology, noise is viewed largely as a nuisance: something entirely
distinct from the processes we are interested in which invariably keeps
getting in the way.  As we have seen in section on origins, such 
characterizations are misleading: noise arises from the very same processes.
While both empirical and statistical literature have frequently emphasized
techniques to remove noise, a central theme of stochastic processes
have emphasized an almost opposite conclusion: noise cannot be subtracted or
averaged away, since the presence of noise can fundamentally alter the
dynamics we observe.
Stochasticity can give rise to a wide array of interesting phenomena
that are often at odds with our intuition. As these examples will
illustrate, such behavior arises through an interaction between
stochastic and non-linear elements of a system, underscoring the
importance of explicitly including stochasticity in ecological
models [@Coulson2004; @Higgins1997].  Importantly, the noise itself
does not have to be large to give rise to these phenomena.  In this
section, I highlight several examples that have shown up frequently
in the literature and prove sufficiently tractable to provide a
simple intuition for the mechanisms involved.

@@

Oscillations arise through the process known as stochastic resonance [@Nisbet1976; @Greenman2003].
The damped oscillations seen in Figure \ref{quasicycles}A are analgous to a child
sitting still on a swing: friction slowly damps the magnitude of
each successive oscillation, while the period of pendulum remains
unchanged.  Driving the pendulum through perturbations timed to
match that natural period can quickly create sustained oscillations
with large magnitudes, but such perfect timing is not required.  In
the model above, white noise effectively drives the pendulum at all
frequencies, though at very low magnitude, since the noise is small
relative to the scale of the dynamics.  However, frequencies that
most closely match the natural period of the damped oscillator are
amplified, resonating with the natural oscillations.  This process
is sufficient to permit even small noise to drive sustained
oscillations in the model indefinitely.  The power spectrum (the
squared norm of the Fourier transform, a measure of what frequencies
are present; see, e.g. @Nisbet1982 for background and examples in
@McKane2005, @Alonso2007 or @Black2012) provides a convenient way to
visualize the effects of these oscillations.  The power spectrum
for a deterministic limit cycle would create a perfect sinusoidal
oscillation and a resulting power spectrum of a delta spike at that
single frequency.  The power spectrum for the quasi-cycle instead
shows a distribution of frequencies, centered at the the resonant
frequency of the damped oscillator [@Nisbet1982].

Since @Nisbet1976, examples of stochastic resonance have been demonstrated in common
models and implicated in empirical examples of increasing complexity, [@Greenman2003],
particularly in higher-dimensional disease and outbreak models [@Blarer1999; @Greeman2003; @McKane2005; @Alonso2007; @Black2010].
@Blarer1999 identified quasi-cycles as a possible
explanation for oscillations and periodic outbreaks in pest ecosystems.
@Bjornstad2004 analyzes the potential for such stochastic effects
to drive cyclic behavior in population abundances of Atlantic bluefin
tuna and Atlantic Cod.  @McKane2005, @Alonso2007, and @Black2010 illustrate
the emergence of quasi-cycles in common disease outbreak models, including
the use of power spectrum analysis.  
@Pineda-Krch2007 compare patterns from stochastically generated cycles to
stable limit cycles they resemble to ask how we can discriminate between
these mechanisms.  Stochastic resonance highlights
one way in which even small noise can play an important role in
driving large scale patterns where we had previously only sought
deterministic explanations. This example also illustrates the
interplay between nonlinear dynamics and stochasticity that is the
hallmark of stochastically driven phenomena.  A stable node with
linear dynamics, such as an Ornstein-Uhlenbeck process, has no
resonant frequency, and could not produce sustained oscillations
(unless the environmental driver itself had a periodic pattern)
[e.g. @Gardiner2009; @Black2010].

@@

<!--Figure 3: Noisy switch -->
```{r noisy_switch, fig.width=7, fig.height=7, fig.cap="Panel A: Data on prevelance of rubella in Copenhagen illustrating potential stochastic switching dynamics between two regimes, reproduced from @Keeling2001. B-D. Stochastic switiching in May's model of alternative stable state dynamics. Panel B: timeseries showning stochastic switching events between regimes. Panel C: separate nonlinear birth and death processes intersect at multiple points creating alternate stable states. Panel D: The corresponding potential well derived from the birth-death model: $U(x) = -\\int b(x) - d(x) \\ ud x $. Parameter values shown r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000, see code in Appendix A. \\label{noisy_switch}"}

@@

Because these transitions between alternative stable states appear as
such a dramatic deviation from the previous pattern (as illustrated
in Figure \ref{noisy_switch}A), it is  tempting to assume they
result from some sudden external force or change in the
environment. As the numerical examples illustrate,
these transitions can be explained entirely by chance
[@Boettiger2012;  @Drake2013].

@@

Transitions from systems with particularly shallow wells have recently
taken on new empirical significance.  Because such transitions become
more likely as the potential well becomes less deep (i.e. the attractor
has become less stable), the onset of such transitions can be interpreted
as an indication of this loss of stability. This phenomenon has been
dubbed "flickering" and suggested as a warning sign that a more
permanent transition to the alternate stable state is approaching [@Brock2010].  
Examples of flickering between alternative
stable states have been suggested in data ranging from geological
timescales at the end of glaciation in the Pleistocene [@Taylor1993]
to ecological scales over a few decades in lake ecosystems [@Wang2012].
Flickering is then our first example of "noise the informer" -- in
which the emergence of a stochastic phenomenon is re-interpreted 
in a way that sheds light on an underlying process.

In examining "noise the creator," we began with a tension between
the interpretation of noise as it is usually presented in the context of
statistical inference: a nuisance to be stripped away, and how it is
viewed by theorist, as a potentially fundamental driver of patterns,
including oscillations, regime shifts, extinction & coexistence.
In this third section, we attempt to re-unite these two largely
divergent views of the role of noise by highlighting how phenomena
from the latter (noisy phenomena) can be used to address the questions
of the former (inferring underlying process from pattern).

Our most provocative example of how noisy phenomena can act to reveal
underlying processes comes from the literature we have just touched upon
in the case of flickering -- a growing set of so-called 'early
warning signs' of critical transitions [e.g. @Scheffer2009].  The fundamental
insight behind all of these signs comes from the exploiting the ability of
equations we encountered in "origins of noise"
to reveal subtle changes in underlying processes before those changes are
visible at the macroscopic (deterministic) level.  This literature focuses
on the phenomenon of a tipping point, in which gradual changes to the environment
show essentially no changes to the behavior of a system (i.e. the abundance of a
population or health of lake ecosystem) followed by a rapid change when the
tipping point is reached that cannot be easily reversed. [@Scheffer2009;
@Scheffer2015].  The fundamental insight of early warning signals has been
to realize that though the average or deterministic dynamics of such a system
do not change significantly prior to the tipping point, the dynamics of the
noise must change in a consistent and predictable manner.

This pattern, known as 'critical slowing down' for reasons that will become 
obvious, was first proposed an an early warning sign for ecological
transitions by @Wissel1984, but has received considerable attention only more recently,
particularly owing to work in empirical systems [@Scheffer2015].
Our analysis in the origins of noise provides sufficient framework to illustrate
this idea completely.  The premise behind the tipping point just described
is that of a fold bifurcation [which I will not detail here, but see @Holling1973; @Scheffer2009; @Strogatz2001].
The key thing we need to know about this bifurcation is that it involves
changes *derivative* of the deterministic skeleton ($\partial_x \left[ b(x) - d(x) \right]_{x=\hat x}$) in
the notation of our example), which can build up for some time without
significant changes to the location of the equilibrium state $\hat x$.
This derivative did not appear in our macroscopic equation, $\tfrac{\mathrm{d}x}{\mathrm{d}t} = b(x) - d(x)$,
but featured regularly in our equations for the noise, such as the 
expected variance, Eq \eqref{var} and auto-correlation, Eq \eqref{autocor}.
This derivative represents the *stability*  of the equilibrium: how quickly
deviations return to the stationary value.  
The magnitude of both variance and auto-correlation must increase
as the stability ($\partial_x \left[ b(x) - d(x) \right]_{x=\hat x}$) decreases
prior to a critical transition.

<!-- Figure 4: Early warning signals -->
```{r tipping, fig.width=7, fig.height=10, fig.cap="Early warning signs of a critical transition in a simulation of increasing serial dilution of budding yeast. Panel A: Data from replicate manipulations by @Dai2015. Panel B: Coefficient of variation for the same experimental data. Panel C: Simulated single time series model of @Dai2015.  Panel D, E: Rolling variance and autocorrelation over simulated data show an increase in advance of the transition.  Simulation details in Appendix A. Warning signal statistics for experimental data not shown, see @Dai2015.\\label{tipping}"}
tipping <- read_csv("../appendixA/tipping.csv")

@@

The trick of the early warning signals approach is to view noise as 
a sequence of countless miniature perturbation experiments.  
Figure \ref{tipping} illustrates early warning indicators in an
experimental manipulation and corresponding simulated model of
@Dai2012 and @Dai2015.  Though early warning signals have been
reported in a growing range of systems [@Scheffer2015], these
laboratory experiments in population dynamics
they observed in serial dilutions of budding yeast _Saccharomyces
cerevisiae_ on a sucrose medium provide the most convincing 
mechanistic interpretation of the underlying fold bifurcation.
Because the relevant proteins are
located in the extra-cellular matrix rather than inside the cells,
budding yeast break down sucrose collaboratively, creating an Allee
effect below which the population is too small to sustain itself.
Increasing the serial dilution rate slowly over time drives the
stable population towards a fold bifurcation which is accompanied
by the patterns of increasing variance and increasing auto-correlation.
This pattern can be seen empirically in lab populations [@Dai2012; @Dai2015]
and explored numerically in a simulation of the yeast dynamics shown in 
Figure \ref{tipping}. As the mechanistic model of yeast population
dynamics is already quite simple, the simulation shown here (code in Appendix B)
corresponds directly to the model proposed by @Dai2015.

@@

In contrast to the examples we have considered previously, work that extends
this analysis into more complex scenarios of structured populations, interacting species,
auto-correlated noise etc are significantly underdeveloped. Issues of spatial
structure have received more attention [e.g. @Guttal2009; @Dakos2011; @Bel2012; @Dai2013]
but still largely idiosyncratic.  Significant challenges remain
in distinguishing these patterns from other causes [@Boettiger2013a].
For instance, as we have seen, variance in a system can increase due to
external drivers without changing the auto-correlation, and both variance
and auto-correlation can increase if a system loses stability without
approaching a saddle-node bifurcation, but will not show flickering.
Teasing these issues apart in greater mathematical formalism and under more
complex and realistic models remain largely open problems.

More broadly, examples of early warning signals such as increasing variance,
auto-correlation, and flickering all point to ways in which stochasticity
and noise-driven phenomena can be transformed from something which obscures
underlying processes to something which reveals them.  I believe this is
role could apply far more broadly than in just the context of stability and
tipping points.  The examples from fold bifurcations already suggest
that stochastic processes could play a similar role in identifying
other critical transitions. For instance, stochastic resonance may be 
visible in advance of a Hopf bifurcation, where a stable state gives
way to stable oscillations (limit cycle) of growing magnitude
[e.g. @Fussmann2000; @Nelson2013; @Strogatz2001].

Interpreting stochastic processes as potential new sources of information
has also proven to be a
rich catalyst for bringing these concepts beyond the realm of abstract
theory. A rich array of experimental and observational tests of early
warning signals [_i.e._ @Carpenter2011; @Dai2012; @Clements2016; @Clements2017;
and many others, _e.g._ see @Scheffer2015 review] has served not merely
to illustrate these phenomena can be seen in laboratory and field ecosystems,
but also to spark suggestions of additional possible indicators and surface
a wide array of practical [@Boettiger2013b; @Scheffer2015], statistical
[e.g. @Boettiger2012d; @Boettiger2012], and theoretical [e.g. @Dai2015;
@Hastings2010] challenges to such application.  This creates a two-way dialog
between empirical and theoretical work that does much to advance the field
as a whole.
